The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-04-28 VegOzone.tex
Updated text with respect to data from SeNorge.

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $7.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring is in coincidence with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} coincides with the estimated onset of the greening season based on the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively (day of the year 110--150 at Svanvik, Appendix Fig~\ref{fig:greening_season_change_Svanvik}).

The density distribution for Svanvik and the respective \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. The spring peak apparently occurs slightly earlier (in the beginning of April). As Svanvik is located at lower altitude than the other sites, the earlier ozone peak may point to a strong interplay between ozone uptake by vegetation after the onset of greening, for at higher altitude plants sprout later. {\bf TODO: Is there a way to confirm this? Actual data for the three sites? Please correct if wrong! Ane? Aud?} The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during night time.

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{greening_season_change_Svanvik}
  \caption{Estimated shift and prolongation of greening season at Svanhovd over the past two decades based on the method and data described in \citet{GMD:Falk2019}.}
  \label{fig:greening_season_change_Svanvik}
\end{figure}
\clearpage

>>>>

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $7.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring is in coincidence with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} begins about one month ahead of the estimated onset of the greening season. Based on data from SeNorge.no and the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively, we find $\left<G_\mathrm{begin}\right> = (150 \pm 14)\,\unit{days of the year}$ at Svanvik (Appendix Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May.

The density distribution for Svanvik and the respective \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. The spring peak apparently occurs slightly earlier (in the beginning of April). As Svanvik is located at lower altitude than the other sites, the earlier ozone peak may point to an interplay with ozone uptake by vegetation. {\bf TODO: The data from SeNorge seems to point to a later onset of greening, than what I interfered from OpenIFS data. This means that the beginning of the growing season is not a good measure for the determination of the ozone spring peak. This does not exclude the early start of pine forest photosynthesis as a driver. Is there a way to confirm this? Actual data for the three sites? Please correct if wrong! Ane? Aud?} The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during night time.

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{greening_season_change_Svanvik}
  \caption{Estimated shift and prolongation of greening season at Svanhovd over the past 6 decades based on data from SeNorge.no {\bf TODO: citation http://www.senorge.no}.}
  \label{fig:greening_season_change_Svanvik}
\end{figure}
\clearpage
